Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-07-26 unimodular-covers.tex
changed statement of theorem 1.6

Oda (\cite{Oda1997}) posed several questions regarding smoothness and the IDP property for lattice polytopes.
%
Following \cite{HaaseHof, Tsuchiya}, we say that a pair $(P, Q)$ of lattice polytopes has the integer decomposition property, or that \emph{the pair $(P,Q)$ is IDP}, if 
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
%
A lattice polytope $Q$ is called \emph{smooth} if it is simple and the primitive edge directions at every vertex form a linear basis for the lattice; equivalently, if the projective toric variety defined by the normal fan of $Q$ is smooth. 
The following versions of Oda's questions are now considered conjectures~\cite{HNPS2008,mfo2007}, and they are open even in dimension three:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item 
\label{itm:smoothIDP}
(Related to problems 2 and 5 in \cite{Oda1997})
Every smooth lattice polytope is IDP.
\item 
\label{itm:mixedIDP}
(Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Every pair $(P,Q)$ of lattice polytopes with $Q$ smooth and the normal fan of $Q$ refining that of $P$ is IDP.
%\begin{align}
%\label{eq:mixedIDP}
%(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
%\end{align}
\end{enumerate}
\end{conjecture}

When the normal fan of a polytope $Q$ refines that of another polytope $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$\paco{did not find this in the google books version of \cite{Grunbaum}; found it in Postnikov-Reiner-Williams, Tsuchiya, a survey by Rekha Thomas, and other places. Perhaps we can leave it wothout reference}. 
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor defined by $P$ on  $X_Q$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}, but observe that what we here call  ``weak Minkowski summand'' is simply called ``Minkowski summand'' there). 
%\paco{added this explanation}
%
%\begin{definition}
%Following \cite{HaaseHof, Tsuchiya} we say that the pair $(P, Q)$ has the integer decomposition property, or \emph{the pair $(P,Q)$ is IDP}, if 
%\begin{align*}
%(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
%\end{align*}
%\end{definition}
% 
%In this language,  \Cref{conj:Oda}\eqref{itm:mixedIDP} can be rephrased as: if $P$ is a weak Minkowski summand of a smooth polytope $Q$ then the pair $(P,Q)$ is IDP.

@@

%\begin{proof}
%Let $P$ be a smooth centrally symmetric lattice $3$-polytope.
%Beck et al.~\cite{BHHHJKM2019} show that $P$ is IDP by covering it with parallelepipeds and unimodular simplices. Once we know parallelepipeds have unimodular covers (\Cref{thm:parallelepipeds}), we conclude $P$ has a unimodular cover too.
%\end{proof}

The two-dimensional case of \Cref{conj:Oda}\eqref{itm:mixedIDP} is known to hold, with three different proofs by Fakhruddin~\cite{Fakhruddin}, Ogata~\cite{Ogata} and Haase et al.~\cite{HNPS2008}. This last one actually shows that smoothness of $Q$ is not needed. In dimension three, however, the conjecture fails without the smoothness assumption. Indeed, if we let $P=Q$ be any non-unimodular \emph{empty tetrahedron}, then $P$ is obviously a weak Minkowski summand of $Q$ but the pair $(P.Q)$ is not IDP. By an empty tetrahedron we mean a lattice tetrahedron containing no lattice points other than its vertices (see the proof of \Cref{lemma:corner} for a classification of them).
\giulia{I didn't like "details on them" and changed it to  "classification of empty tetrahedra". Is this ok?}
\paco{ok. But left it as "classification of them"}

An alternative approach to \Cref{conj:Oda}\eqref{itm:mixedIDP} is via Cayley sums, which we discuss in  \Cref{sec:cayley}. 
Recall that the \emph{Cayley sum} of two lattice polytopes $Q_1,Q_2\subset \R^d$ is the lattice polytope
\[
\cayley(Q_1,Q_2) := \conv(Q_1\times\{0\} \cup Q_2 \times \{1\}) \subset \R^3.
\]
We normally require $\cayley(P,Q)$ to be full-dimensional (otherwise we can delete coordinates) but this does not need $Q_1$ and $Q_2$ to be full-dimensional. It only requires the linear subspaces parallel to them to span $\R^d$.

@@

\begin{theorem}
\label{thm:cayley}
Let $P$ and $Q$ be lattice polygons, with $P$ a weak Minkowski summand of $Q$. Then the Cayley sum $\cayley(P,Q)$ has a unimodular cover.
\end{theorem}

@@

See \Cref{fig:corner}. Observe that $u+v=p_1+p_2=q_3+q_4$.
Now, this implies that the quadrilateral $\conv(p_1q_4p_2q_3)$ contains a fundamental domain for the lattice $\Z^2\times\{0\}$. Hence, its translate $\conv(q_2p_3q_1p_4)$ contains a fundamental domain for $\Z^2\times\{1\}$ and, in particular, it contains at least one lattice point other than $p_3$ and $p_4$. By central symmetry around its center $\left(\frac{a}2,-\frac{b}2,1\right)$, $\conv(q_2p_3q_1p_4)$ must contain lattice points in both triangles $\conv(q_2p_3p_4)\subset T_1$ and $\conv(q_1p_3p_4)\subset T_2$.
%
\giulia{ditto}
\paco{not sure about this one. It looks like the red dots lie on edges of $C(T)$}
\begin{figure}[htb]
\includegraphics[scale=.25]{Whites}
\caption{The parallelograms $\conv(p_1p_2q_3q_4)$ and $\conv(q_2q_1p_4p_3)$. The large red dots are lattice points.}
\label{fig:corner}
\end{figure}
%\paco{Alternative arguments: one can explicitly compute coordinates of lattice points in $T_1$ and $T_2$, or one can say that the isomorphism $T_{a,b} \cong T_{a^{-1},b}$ of White tetrahedra exchanges the edges at $\{z=0\}$ and at $\{z=1\}$.}
\end{proof}

@@

\section{ Cayley sums}
\label{sec:cayley}
Let $P$ and $Q$ be two lattice polytopes in $\R^d$. We do not require them to be full-dimensional, but we assume their Minkowski sum is. Remember that the \emph{Minkowski sum} $P+Q$ and the \emph{Cayley sum} of $P$ and $Q$ are defined as:
\begin{gather*}
P + Q := \{ p+q \in \R^d: p\in P, q \in Q\}.\\
\cayley (P,Q) = \conv( P\times\{0\} \cup Q\times \{1\}) \in \R^{d+1}.
\end{gather*}

@@

%\giulia{added picture, we may not want it}
%\paco{looks ok}
\begin{figure}[htb]
\scalebox{.75}{\input{strips.pdf_t}}
\caption{The strips  of Lemma \ref{lemma:cayley}}
\label{fig:strips}
\end{figure}

\begin{corollary}
\label{coro:covercayley}
Let $T$ be an empty lattice tetrahedron contained in the Cayley sum $\cayley(P,Q)$, where $P$ is a weak Minkowski summand of $Q$. Then, $T$ can be covered by unimodular tetrahedra contained in $\cayley(P,Q)$.
\end{corollary}

@@

We let $H_1=\{f_q(x) \leq 0\}$ and $H_2=\{f_q(x) \geq 0\}$; similarly let $V_1=\{f_p(x) \leq 0\}$ and $V_2=\{f_p(x) \geq 0\}$. 
\giulia{we need to make these closed, so that the case $P$ one dimensional fits into this same proof}\paco{ok}
In the figures, we draw $p$ as a vertical segment and $q$ as a horizontal one, so that $H_i \cap V_j$ are the four quadrants. 
See \Cref{fig:setup}.
%
\begin{figure}[htb]
\includegraphics[scale=.3]{setup.png}
\caption{Setup for the proof of \Cref{lemma:cayley}}
\label{fig:setup}
\end{figure}

@@

\bibitem{Grunbaum}
{Branko Gr\"unbaum}.
{\em {C}onvex {P}olytopes}.
Wiley, London, 1967.

>>>>

Oda (\cite{Oda1997}) posed several questions regarding smoothness and the IDP property for lattice polytopes.
%
Following \cite{HaaseHof, Tsuchiya}, we say that a pair $(P, Q)$ of lattice polytopes has the integer decomposition property, or that \emph{the pair $(P,Q)$ is IDP}, if 
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
%
A lattice polytope $Q$ is called \emph{smooth} if it is simple and the primitive edge directions at every vertex form a linear basis for the lattice; equivalently, if the projective toric variety defined by the normal fan of $Q$ is smooth. 
The following versions of Oda's questions are now considered conjectures~\cite{HNPS2008,mfo2007}, and they are open even in dimension three:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item 
\label{itm:smoothIDP}
(Related to problems 2 and 5 in \cite{Oda1997})
Every smooth lattice polytope is IDP.
\item 
\label{itm:mixedIDP}
(Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Every pair $(P,Q)$ of lattice polytopes with $Q$ smooth and the normal fan of $Q$ refining that of $P$ is IDP.
\end{enumerate}
\end{conjecture}

When the normal fan of a polytope $Q$ refines that of another polytope $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is easily seen to be equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$.
%\paco{did not find this in the google books version of \cite{Grunbaum}; found it in Postnikov-Reiner-Williams, Tsuchiya, a survey by Rekha Thomas, and other places. Perhaps we can leave it wothout reference} 
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor defined by $P$ on  $X_Q$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}, but observe that what we here call  ``weak Minkowski summand'' is simply called ``Minkowski summand'' there).

@@

The two-dimensional case of \Cref{conj:Oda}\eqref{itm:mixedIDP} is known to hold, with three different proofs by Fakhruddin~\cite{Fakhruddin}, Ogata~\cite{Ogata} and Haase et al.~\cite{HNPS2008}. This last one actually shows that smoothness of $Q$ is not needed. In dimension three, however, the conjecture fails without the smoothness assumption. Indeed, if we let $P=Q$ be any non-unimodular \emph{empty tetrahedron}, then $P$ is obviously a weak Minkowski summand of $Q$ but the pair $(P.Q)$ is not IDP. By an empty tetrahedron we mean a lattice tetrahedron containing no lattice points other than its vertices (see the proof of \Cref{lemma:corner} for a classification of these).
\giulia{I didn't like "details on them" and changed it to  "classification of empty tetrahedra". Is this ok?}
\paco{ok. But left it as "classification of them"}
\giulia{is it of them or of these!? my grammar needs some refresh}

An alternative approach to \Cref{conj:Oda}\eqref{itm:mixedIDP} is via Cayley sums, which we discuss in  \Cref{sec:cayley}. 
Recall that the \emph{Cayley sum} of two lattice polytopes $P,Q\subset \R^d$ is the lattice polytope
\[
\cayley(P,Q) := \conv(P\times\{0\} \cup Q \times \{1\}) \subset \R^3.
\]
We normally require $\cayley(P,Q)$ to be full-dimensional (otherwise we can delete coordinates) but this does not need $P$ and $Q$ to be full-dimensional. It only requires the linear subspaces parallel to them to span $\R^d$.

@@

\begin{theorem}
\label{thm:cayley}
Let $Q$ be lattice polygon, and $P$ a weak Minkowski summand of $Q$. Then the Cayley sum $\cayley(P,Q)$ has a unimodular cover.
\end{theorem}\giulia{changed this so that only $Q$ need be a lattice polygon}

@@

See \Cref{fig:corner}. Observe that $u+v=p_1+p_2=q_3+q_4$.
Now, this implies that the quadrilateral $\conv(p_1q_4p_2q_3)$ contains a fundamental domain for the lattice $\Z^2\times\{0\}$. Hence, its translate $\conv(q_2p_3q_1p_4)$ contains a fundamental domain for $\Z^2\times\{1\}$ and, in particular, it contains at least one lattice point other than $p_3$ and $p_4$. By central symmetry around its center $\left(\frac{a}2,-\frac{b}2,1\right)$, $\conv(q_2p_3q_1p_4)$ must contain lattice points in both triangles $\conv(q_2p_3p_4)\subset T_1$ and $\conv(q_1p_3p_4)\subset T_2$.
%

%\begin{figure}[htb]
%\includegraphics[scale=.28]{Whites}
%\caption{The parallelograms $\conv(p_1p_2q_3q_4)$ and $\conv(q_2q_1p_4p_3)$. The large red dots are lattice points.}
%\label{fig:corner}
%\end{figure}

\end{proof}

@@

\section{ Cayley sums}
\label{sec:cayley}
Let $P$ and $Q$ be two lattice polytopes in $\R^d$. We do not require them to be full-dimensional, but we assume their Minkowski sum is. Remember that the \emph{Minkowski sum} $P+Q$ and the \emph{Cayley sum} of $P$ and $Q$ are defined as:
\begin{gather*}
P + Q := \{ p+q \in \R^d: p\in P, q \in Q\} \subset \R^d,\\
\cayley (P,Q) = \conv( P\times\{0\} \cup Q\times \{1\}) \subset \R^{d+1}.
\end{gather*}

@@

\begin{figure}[htb]
\scalebox{.75}{\input{strips.pdf_t}}
\caption{The strips  of Lemma \ref{lemma:cayley}}
\label{fig:strips}
\end{figure}

\begin{corollary}
\label{coro:covercayley}
Let $T$ be an empty lattice tetrahedron contained in the Cayley sum $\cayley(P,Q)$, where $Q$ is a lattice polygon and $P$ is a weak Minkowski summand of $Q$. Then, $T$ can be covered by unimodular tetrahedra contained in $\cayley(P,Q)$.
\end{corollary}

@@

We let $H_1=\{f_q(x) \leq 0\}$ and $H_2=\{f_q(x) \geq 0\}$; similarly let $V_1=\{f_p(x) \leq 0\}$ and $V_2=\{f_p(x) \geq 0\}$. 
%\giulia{we need to make these closed, so that the case $P$ one dimensional fits into this same proof}\paco{ok}
In the figures, we draw $p$ as a vertical segment and $q$ as a horizontal one, so that $H_i \cap V_j$ are the four quadrants. 
See \Cref{fig:setup}.
%
\begin{figure}[htb]
\includegraphics[scale=.3]{setup.png}
\caption{Setup for the proof of \Cref{lemma:cayley}}
\label{fig:setup}
\end{figure}

@@

%\bibitem{Grunbaum}
%{Branko Gr\"unbaum}.
%{\em {C}onvex {P}olytopes}.
%Wiley, London, 1967.
